Mental simulation as Bayesian Quadrature

### 2013-05-31 man/nips-2013.tex
Polished results section

% I think it would be compelling to add a second row to Equation 5
% where you show the averages and s.e. at each rotation for many noisy
% runs of the algorithm (possibly on different stimuli). If you can
% show that the average is a linear function, in a way that looks like
% the classic behavioral data, that will be compelling.

\begin{figure}[t]
  \centering
  \includegraphics[width=\textwidth]{../figures/model_rotations.pdf}
  \caption{\textbf{Model rotations.} Each subplot shows the
    correspondence between the true angle of rotation ($R$) for $h_1$
    stimuli and the amount of rotation performed by the model. Dotted
    lines indicate a perfect 1:1 relationship. The \naive{} and
    parametric models (left and center panels, respectively) are
    identical; they show modest agreement until roughly
    $\frac{\pi}{2}$, after which agreement decreases as they tend to
    encounter local maxima. The nonparametric model (right panel)
    exhibits a significant correlation between the true angle and the
    amount the model rotates.}
  \label{fig:rotations}
\end{figure}

We evaluated each model's performance on 20 randomly generated shapes
which had between 3 and 6 vertices, inclusive, (e.g. $X_a$ in Figure
\ref{fig:stimuli}). For each shape, we computed 18 ``different'' and
18 ``same'' stimuli pairs, with $R$ spaced at $20^\prime$ increments
between 0 and 360. ``Different'' pairs were created by reflecting
$X_a$ across the $y$-axis and rotating it by $R$; for ``same'' pairs,
$X_a$ was rotated by $R$. \TODO{need to mention that shapes were
  chosen to be discriminable by the gold standard}

We considered three metrics of performance in particular:
\begin{itemize}
\item \textit{Responses}: how accurate is the model at choosing the
  correct hypothesis? This is defined to be the mean error ($\ME{}$),
  or fraction of times the model picks the incorrect hypothesis.
\item \textit{Rotations}: for those ``same'' pairs which the model
  gets correct, how correlated are the true angles of rotation with
  the rotations performed by the model?  This is defined to be the
  Pearson's correlation coefficient $\rho$ between $R$ and $\vert
  \mathbf{R}\vert$. Figure \ref{fig:rotations} shows individual points
  corresponding to true rotation vs. rotation by the model, for each
  model.
\item \textit{Estimates of $Z$}: how accurate is the model's estimate
  of $Z$? This is defined to be the mean squared error ($\MSE{}$)
  between the model's estimate of $Z$ and the ``gold standard'' value,
  where the error has scale such that $\MSE{}=0$ indicates no error,
  and $\MSE{}=1$ indicates maximum error. Figure \ref{fig:accuracy}
  shows plots of the true (gold standard) value of $Z$ vs. the model's
  estimate, for each model.
\end{itemize}

The mean error of the gold standard model was $\GSME{}$, indicating
that $Z$ is a sufficient approximation of $\int_R p(X_b\ \vert\
X_R)p(R)\ \mathrm{d}R$ for these particular stimuli.

@@

The \naive{} model's response error rate was better than chance with
$\NaiveME{}$. Closer inspection reveals that much of this comes from
$h_0$ stimuli ($\NaiveMEdiff{}$), where the asymmetric linear
interpolation gives an overestimate of $Z$ (e.g., in Figure
\ref{fig:li}). There is some error from $h_1$ stimuli as well
($\NaiveMEsame{}$), which happens when the hill climbing search finds
a local maxima and terminates prematurely, thus resulting in a low
estimate of $Z$.

The correlation between the average number of rotations computed by
the \naive{} model and the true angle of rotation was
$\Naivecorr{}$. The shape of the data is more interesting (see Figure
\ref{fig:rotations}, top left panel): the \naive{} model actually
corresponds extremeley well to the true angle of rotation for
$R<\frac{\pi}{2}$ ($\Naivecorrsm{}$). This is unsurprising, because
the closer the true angle is to zero, the less the model has to
rotate, and the less likely it will get stuck on local maxima. Thus,
it is more likely to locate the global maximum, which corresponds to
the true angle of rotation. For $R>\frac{\pi}{2}$, we see an
increasing tendency to under-rotate due to getting stuck on local
maxima, as well as a tendency to over-rotate if the wrong rotation
direction was chosen.

The \naive{} model's error on $h_0$ stimuli is reflected in its
accuracy in estimating $Z$ ($\NaiveMSE{}$). Figure \ref{fig:accuracy}
(left panel) shows individual estimates for each stimulus, color-coded
by true hypothesis. The model overestimates $Z$ for most of the $h_0$
and $h_1$ stimuli, and underestimates several of the $h_1$ stimuli as
well.

@@

The parametric model is also fairly inaccurate ($\VMME{}$), making
more errors on $h_1$ stimuli ($\VMMEsame{}$) than $h_0$ stimuli
($\VMMEdiff{}$). This is because the Von Mises distribution only has a
single peak and thus will underestimate $Z$ for multimodal similarity
functions. However, it does more accurately estimate $Z$ for stimuli
that the \naive{} model would overestimate ($\VMMSE{}$). This
underestimation is severe enough that in most cases where it gets
stuck on a local maxima, it produces the incorrect judgment. So, the
correlation for \textit{correct} responses is actually improved
($\VMcorr{}$) and would be near perfect if the correct diretion of
rotation was initially chosen (see Figure \ref{fig:rotations}, top
center).

@@

The nonparametric model is much more accurate in choosing the correct
hypothesis ($\BQME{}$), erring approximately equally for both types of
stimuli. We note that this is very close to the 3.2\% error rate
reported by \cite{Shepard1971}, though further experimentation is
necessary to determine whether the types of errors people make align
with those of the model.

The average rotations computed by the nonparametric model were
strongly correlated with the true rotations ($\BQcorr$, see Figure
\ref{fig:rotations}, bottom right). This is because the nonparametric
model does not get stuck as easily on local optima: it will continue
rotating until it is confident that its estimate of $Z$ is
accurate. Indeed, its active attempts to obtain an accurate estimate
of $Z$ are largely succeessful ($\BQMSE{}$). Moreover, it has the
capacity to ``restart'' and can recover from initially rotating in the
incorrect direction, as illustrated by the fewer number of extreme
over-rotations in Figure \ref{fig:rotations} (top right).

\begin{figure}[t]
  \centering
  \includegraphics[width=\textwidth]{../figures/Z_accuracy.pdf}
  \caption{\textbf{Accuracy in estimating $Z$.} Each subplot shows the
    true (``gold standard'') value of $Z$ vs. the value estimated by
    the model. Black dotted lines indicate a perfect 1:1
    correspondence between the true and estimated values. All models
    perform well for $h_0$ stimuli, but exhibit different behavior for
    $h_1$ stimuli. The \naive{} model (left panel) tends to either
    greatly overestimate, or underestimate. The parametric model
    (center panel) is very accurate for about half the $h_1$ stimuli,
    and severely underestimates the rest. The nonparametric model
    (right panel) maintains a decent correspondence with the true $Z$,
    with the exception of a handful of outliers.}
  \label{fig:accuracy}
\end{figure}

>>>>

\begin{figure}[t]
  \centering
  \includegraphics[width=\textwidth]{../figures/model_rotations.pdf}
  \caption{\textbf{Model rotations.} Top: each subplot shows the
    correspondence between the true angle of rotation ($R$) for
    ``same'' pairs and the amount of rotation performed by the
    model. Bottom: each subplot shows the models' mean rotations over
    stimuli pairs that were judged correctly. Black dots correspond to
    ``same'' pairs, and blue lines correspond to ``different'' pairs
    (for which the mean rotation is constant across true rotations, as
    the true rotation is undefined). Error bars/shaded regions
    indicate one standard deviation, and the dotted lines indicate the
    least-squares fit to the ``same'' pairs.}
  \label{fig:rotations}
\end{figure}

We evaluated each model's performance on 20 shapes which had between 3
and 6 vertices, inclusive, (e.g. $X_a$ in Figure
\ref{fig:stimuli}). For each shape, we computed 18 ``same'' and 18
``different'' stimuli pairs, with $R$ spaced at $20^\prime$ increments
between 0 and 360, as in \cite{Shepard1971}. ``Same'' pairs were
created by rotating $X_a$ by $R$; the same was true for ``different'',
except that $X_a$ was also reflected across the $y$-axis. Each shape
was generated randomly, but under the constraint that the
``same''/``different'' pairs created from the shape could both be
judged according to the ``gold standard''.

We considered three metrics of performance in particular:
\begin{itemize}
\item \textit{Responses}: How accurate was the model at choosing the
  correct hypothesis? This was defined to be the mean error ($\ME{}$),
  or fraction of times the model picks the incorrect hypothesis.
\item \textit{Rotations}: For those ``same'' pairs which the model
  judged correct, how correlated were the model's rotations with the
  true angles of rotation?  We quantified this using the Pearson's
  correlation coefficient $\rho$ for $R$ vs. $\vert
  \mathbf{R}\vert$. The top row in Figure \ref{fig:rotations} shows
  individual points corresponding to true rotations vs. rotations by
  the models for all stimuli, including those judged incorrectly. The
  bottom row shows mean model rotations, across only those stimuli
  which were judged correctly. We structured the analysis in this way
  to better compare to the results of \cite{Shepard1971}, which also
  excluded incorrectly judged stimuli.
\item \textit{Estimates of $Z$}: How accurate was the model's estimate
  of $Z$? We defined this quantity as the mean squared error
  ($\MSE{}$) between the model's estimate of $Z$ and the ``gold
  standard'' value, where the error has scale such that $\MSE{}=0$
  indicates no error and $\MSE{}=1$ indicates maximum error. Figure
  \ref{fig:accuracy} shows plots of the true (``gold standard'') value
  of $Z$ vs. the model's estimate, for each model.
\end{itemize}

@@

The \naive{} model's response error rate was better than chance, with
$\NaiveME{}$. Closer inspection reveals that much of this comes from
``different'' pairs ($\NaiveMEdiff{}$), where the asymmetric linear
interpolation may give an overestimate of $Z$ (e.g. Figure
\ref{fig:li}).  This intuition is confirmed by the model's accuracy
($\NaiveMSE{}$) in estimating $Z$: as shown in the left panel of
Figure \ref{fig:accuracy}, it overestimated for nearly all stimuli
pairs.

The correlation between the \naive{} model's average rotation and the
true angle of rotation was $\Naivecorr{}$ (Figure \ref{fig:rotations},
bottom left). More complex patterns are revealed by examining the
shape of the raw data in Figure \ref{fig:rotations}, top left: the
\naive{} model corresponds extremeley well to the true angle of
rotation for approximately $R<\frac{\pi}{2}$. This is unsurprising,
because the closer the true angle is to zero, the less the model has
to rotate, and the less likely it will get stuck on local
maxima. Thus, it is more likely to locate the global maximum, which
corresponds to the true angle of rotation. For $R>\frac{\pi}{2}$, we
see an increasing tendency to under-rotate due to getting stuck on
local maxima, as well as a tendency to over-rotate if the wrong
direction was initially chosen.

@@

The parametric model's error in determining whether the objects were
identical was $\VMME{}$, making more errors on ``same'' pairs
($\VMMEsame{}$) than ``different'' pairs ($\VMMEdiff{}$). Rather than
overestimating as with the \naive{} model, however, the parametric
model largely underestimated $Z$. This is unsurprising: because the
Von Mises distribution only has a single peak, it necessarily
underestimates $Z$ for multimodal similarity functions. Overall, the
parametric model had an error rate of $\VMMSE{}$ in estimating $Z$,
which was significantly more accurate than that of the \naive{} model.

The correlation between the parametric model's average rotation and
the true angle of rotation was also higher ($\VMcorr{}$), though this
is because more of the underestimated stimuli pairs were excluded from
the analysis. The individual rotations performed by the parametric and
\naive{} models were actually identical (see Figure
\ref{fig:rotations}, upper left and center) because both models use
the same hill-climbing stopping strategy.

\begin{figure}[t]
  \centering
  \includegraphics[width=\textwidth]{../figures/Z_accuracy.pdf}
  \caption{\textbf{Accuracy in estimating $Z$.} Each subplot shows the
    true (``gold standard'') value of $Z$ vs. the value estimated by
    the model. Black dotted lines indicate a perfect 1:1
    correspondence between the true and estimated values. The \naive{}
    model (left panel) tends to overestimate. The parametric model
    (center panel) is very accurate for about half the ``same'' pairs,
    and severely underestimates the rest. The nonparametric model
    (right panel) maintains a decent correspondence with the true
    $Z$.}
  \label{fig:accuracy}
\end{figure}

@@

The nonparametric model was much more accurate in choosing the correct
hypothesis than the other two models ($\BQME{}$). We note that this is
very close to the 3.2\% error rate reported by \cite{Shepard1971},
though further experimentation is necessary to determine whether the
types of errors people make align with those of the model.

The average rotations computed by the nonparametric model were
strongly correlated with the true rotations ($\BQcorr$, see Figure
\ref{fig:rotations}, bottom right). Because the nonparametric model
has the capacity to ``restart'', it could recover from rotating in the
incorrect direction (e.g. Figure \ref{fig:bq}) and thus did not
over-rotate as frequently.  It also under-rotated less frequently:
because the model continues to rotate until it is confident that its
estimate of $Z$ is accurate, it does not get stuck as easily on local
optima.  Indeed, its active attempts to obtain an accurate estimate of
$Z$ were largely succeessful, with $\BQMSE{}$.
